Spin-orbit-coupling-induced squeezing in the optical lattice clock

### 2018-07-19 soc_squeezing/soc_squeezing_notes.tex
made notes cover the continuous-drive protocol before the pulsed one, and added section on benchmarkin the OAT and TAT Hamiltonians against FH simulations

\section{Effective spin squeezing}
\label{sec:OAT}

In total, free evolution of atoms on a lattice is governed by the
Hamiltonian
\begin{align}
  H_{\t{free}}
  = H_{\t{int}} + H_{\t{SOC}}
  = -\f{U}{L} \v S\c\v S - \sum_n h_n s_n^z,
\end{align}
for $n$ indexing an individual atom.  When the ``local'' fields $h_n$
are small in magnitude compared to the collective spin gap $\eta U$
and we initialize all atoms in the ground-state subspace of
$H_{\t{int}}$, we can treat the action of the spin-orbit coupling
Hamiltonian perturbatively, which yields the effective free-evolution
Hamiltonian (see Appendix \ref{sec:squeezing_derivation} or Peiru's
notes)
\begin{align}
  H_{\t{free,eff}} = -\bar h S_z + \chi S_z^2,
\end{align}
where
\begin{align}
  \bar h \equiv \f1N \sum_n h_n,
  &&
  \tilde h^2 \equiv \f1N \sum_n \p{h_n - \bar h}^2,
  &&
  \chi \equiv \f{\tilde h^2}{\p{N-1}\eta U}.
\end{align}
At temperatures $T\gg J_0$, the mean effective field $\bar h=0$, while
for $T\sim J_0$ the collective rotation due to the $S_z$ term can be
eliminated by use of a rotating frame.  The free-evolution Hamiltonian
thus simplifies further to the one-axis twisting Hamiltonian
\begin{align}
  H_{\t{OAT}}^z = \chi S_z^2.
  \label{eq:H_OAT}
\end{align}
An example set of parameters and values for one-axis twisting via
$H_{\t{OAT}}^z$ with strontium-87 in a 1-D magic-wavelength lattice is
provided in Table \ref{tab:parameters}, along with the optimal
one-axis twisting squeezing time
$t_{\t{opt}}^{\t{OAT}}\sim N^{-2/3}\chi^{-1}\sim\eta
N^{1/3}=N^{4/3}/L$.  Lattice depths are provided in units of the
lattice recoil energy $E_R$.

@@

@@

While one-axis twisting provides a spin fluctuation noise floor which
scales with atom number as $\sim1/N^{2/3}$, it is possible to beat
this limit and achieve a noise floor scaling as $\sim1/N$ via
so-called two-axis twisting.  We can transform the one-axis twisting
Hamitlonian in \eqref{eq:H_OAT} into an effective two-axis twisting
Hamiltonian by using a pulsed drive sequence developed in
ref.~\cite{liu2011spin}.  The necessary drive sequence requires the
pulsed application of a clock laser, which turns on a collective spin
Hamiltonian
\begin{align}
  H_{\t{clock}} = -\Omega \sum_n s_n^x = -\Omega S_x.
\end{align}
If $\abs{\Omega}\gg N\chi$, then we can apply $H_{\t{clock}}$ for
short times $\theta/\abs{\Omega}\ll\p{N\chi}^{-1}$ (during which we
can neglect $H_{\t{eff}}$) to realize collective rotations of the form
$\exp\p{\pm i\theta S_x}$.  Thus
\begin{enumerate*}[label=(\roman*)]
\item acting with $\exp\sp{i\p{\pi/2}S_x}$,
\item waiting for a time $2\tau$,
\item acting with $\exp\sp{-i\p{\pi/2}S_x}$, and
\item waiting for a time $\tau$
\end{enumerate*}
realizes the unitary
\begin{align}
  U_1 = \exp\p{-i\tau\chi S_z^2} \exp\p{-i\f{\pi}{2}S_x}
  \exp\p{-i2\tau\chi S_z^2} \exp\p{i\f{\pi}{2}S_x}
  = \exp\p{-i\tau\chi S_z^2} \exp\p{-i2\tau\chi S_y^2}.
\end{align}
If $\tau\chi\ll1$, then
\begin{align}
  U_1 \approx \exp\sp{-i\tau\chi\p{S_z^2 + 2S_y^2}}
  = \exp\sp{-i\tau\chi\p{S^2 + S_y^2 - S_x^2}}
  \simeq \exp\sp{-i\tau\chi\p{S_y^2 - S_x^2}},
\end{align}
where $\simeq$ denotes equality up to an overall phase, which holds
above because $S^2$ is conserved by all dynamics.  Repeating the above
sequence $n$ times for a time $t=3n\tau$ thus results in the unitary
\begin{align}
  U = \exp\sp{-in\tau\chi\p{S_y^2 - S_x^2}}
  = \exp\sp{-i\p{t/3}\chi\p{S_y^2 - S_x^2}},
\end{align}
which is equivalently generated by evolution for a time $t$ under the
two-axis countertwisting Hamiltonian
\begin{align}
  H_{\t{TAT}}^{y,x} = \f{\chi}{3}\p{S_y^2 - S_x^2}.
  \label{eq:H_TAT_pulse}
\end{align}
Despite the factor of 1/3 in the effective two-axis countertwisting
strength, the optimal squeezing time for this two-axis twisting
Hamiltonian is generally close to that for one-axis twisting with
$H_{\t{OAT}}^z$ in \eqref{eq:H_OAT} \cite{liu2011spin}.

\note{[MP: In practice, of course, the gates $\exp\p{i\theta S_x}$
  take a nonzero amount of time to implement.  For the purposes of
  simulation, we can probably approximate these gates by square pulses
  centered on the appropriate time.  I can do an error analysis on
  this sequence to determine the effect of the $\O\p{\tau^2\chi^2}$
  terms in the exponent of $U_1$, but it is probably simpler to just
  try this out in simulations.]}

\note{[MP: I expect this pulsed-drive protocol to be worse than the
  continuous-drive protocol below.  If we decide that we are not
  interested in the pulsed drive, we can merely mention it and throw
  in a citation of ref.~\cite{liu2011spin} without devoting an entire
  section to the subject.]}

\section{Two-axis twisting: continuous drive}

Rather than pulsing the clock laser, we can apply a continuous drive
to realize the Hamiltonian
\begin{align}
  H_{\t{drive}} = H_{\t{free}} + H_{\t{clock}}
  = -\f{U}{L} \v S\c\v S - \bar h S_z - \Omega S_x - \sum_n b_n s_n^z,
\end{align}
where the drive strength $\Omega$ may generally depend on time.  As
shown in Appendix \ref{sec:effective_squeezing_drive}, the addition of
a drive with $\abs{\Omega}\ll\eta\abs{U}=N\abs{U}/L$ does not modify
the squeezing behavior of the spin-orbit coupling terms in
$H_{\t{drive}}$, whose effective action on the $S=N/2$ manifold is
captured by
\begin{align}
  H_{\t{drive,eff}} = -\bar h S_z - \Omega S_x + \chi S_z^2.
  \label{eq:H_drive_eff}
\end{align}
To follow the continuous-drive two-axis twisting prescription in
ref.~\cite{huang2015twoaxis}, we now
\begin{enumerate*}[label=(\roman*)]
\item blue-detune the clock laser by $\bar h$ and shift into an
  appropriate rotating frame to eliminate the $\sim S_z$ term in
  \eqref{eq:H_drive_eff},
\item modulate the drive as $\Omega\p{t}=\beta\omega\cos\p{\omega t}$
  with $\omega\gg N\chi$, and
\item move into the rotating frame of $-\Omega\p{t}S_x$.
\end{enumerate*}
After a secular approximation which relies on $\omega\gg N\chi$, this
procedure results in the effective Hamiltonian
\begin{align}
  H_{\t{drive,eff}}^{\t{mod}}
  = \f{\chi}{2} \p{\sp{\J_0\p{2\beta}+1} S_z^2
    - \sp{\J_0\p{2\beta}-1} S_y^2},
\end{align}
where $\J_0$ is the zero-order Bessel function of the first kind.
Choosing a modulation index $\beta$ such that $\J_0\p{2\beta}=1/3$ or
$\J_0\p{2\beta}=-1/3$ then respectively results in the two-axis
twisting Hamiltonians
\begin{align}
  H_{\t{TAT}}^{z,x}
  = \f{\chi}{3} \p{2 S_z^2 + S_y^2}
  \simeq \f{\chi}{3} \p{S_z^2 - S_x^2},
  &&
  H_{\t{TAT}}^{y,x}
  = \f{\chi}{3} \p{S_z^2 + 2 S_y^2}
  \simeq \f{\chi}{3}\p{S_y^2 - S_x^2},
  \label{eq:H_TAT_drive}
\end{align}
where $\simeq$ denotes equality up to a global energy shift. Given the
set of parameters in Table \ref{tab:parameters}, the two-axis twisting
Hamiltonian $H_{\t{TAT}}^{z,x}$ can be achieved with a drive frequency
$\omega=N\abs{\chi U/L}^{1/2}\approx27\times2\pi$ Hz and modulation
index $\beta=0.906$.  The optimal squeezing time for this protocol is
$t_{\t{opt}}^{\t{TAT}}\approx0.82$ seconds.  A comparison of squeezing
through $H_{\t{OAT}}^z$ and $H_{\t{TAT}}^{z,x}$ for the initial state
$\ket{Y}\equiv\bigotimes_n\p{\ket{\g}_n+i\ket{\e}_n}/\sqrt2$ via the
parameters in Table \ref{tab:parameters} is provided in Figure
\ref{fig:squeezing_comparison}.  The figure of merit for squeezing is
the normalized minimal variance of spin in the plane orthogonal to the
mean spin vector $\bk{\v S}$:
\begin{align}
  \xi^2 \equiv \f{N}{\abs{\bk{\v S}}^2}
  \min_{\hat{\v n}\perp\bk{\v S}} \bk{\p{\v S\c\hat{\v n}}^2}
\end{align}
for $\abs{\hat{\v n}}=1$.

\begin{figure}
  \centering \includegraphics{squeezing_comparison_TAT.pdf}
  \caption{Comparison of squeezing through $H_{\t{OAT}}^z$ and
    $H_{\t{TAT}}^{z,x}$ via the parameters in Table
    \ref{tab:parameters} and the continuous two-axis twisting protocol
    with $\omega=27\times2\pi$ Hz and $\beta=0.906$.}
  \label{fig:squeezing_comparison}
\end{figure}

\section{Discussions}

\note{[MP: We should clarify in the notes what exactly the conditions
  are for the effective Hamiltonians we derive, i.e.~the spin model
  and the one-axis twisting Hamiltonian.]}

\note{[MP] Future ideas:} Make use of a cavity to enhance the one-axis
twisting strength \cite{hu2017vacuum}?

\appendix

>>>>

\section{Spin squeezing and one-axis twisting}
\label{sec:OAT}

In total, free evolution of atoms on a lattice is governed by the
Hamiltonian
\begin{align}
  H_{\t{free}}
  = H_{\t{int}} + H_{\t{SOC}}
  = -\f{U}{L} \v S\c\v S - \sum_n h_n s_n^z,
\end{align}
for $n$ indexing an individual atom.  When the ``local'' fields $h_n$
are small in magnitude compared to the collective spin gap $\eta U$
and we initialize all atoms in the ground-state subspace of
$H_{\t{int}}$, we can treat the action of the spin-orbit coupling
Hamiltonian perturbatively, which yields the effective free-evolution
Hamiltonian (see Appendix \ref{sec:squeezing_derivation} or Peiru's
notes)
\begin{align}
  H_{\t{free,eff}} = -\bar h S_z + \chi S_z^2,
\end{align}
where
\begin{align}
  \bar h \equiv \f1N \sum_n h_n,
  &&
  \tilde h^2 \equiv \f1N \sum_n \p{h_n - \bar h}^2,
  &&
  \chi \equiv \f{\tilde h^2}{\p{N-1}\eta U}.
\end{align}
At temperatures $T\gg J_0$, the mean effective field $\bar h=0$, while
for $T\sim J_0$ the collective rotation due to the $S_z$ term can be
eliminated by use of a rotating frame.  The free-evolution Hamiltonian
thus simplifies further to the one-axis twisting Hamiltonian
\begin{align}
  H_{\t{OAT}}^z = \chi S_z^2,
  \label{eq:H_OAT}
\end{align}
which generates spin squeezing dynamics with metrological
applications.  Specifically, one-axis twisting allows for measurements
of collective spin in the plane orthogonal to the mean spin vector
$\bk{\v S}$ with a spin fluctuation noise floor which scales as
$\sim1/N^{2/3}$ (additional info in Peiru's notes\note{(?)}).

An example set of parameters for realizing one-axis twisting via
$H_{\t{OAT}}^z$ with strontium-87 in a 1-D magic-wavelength lattice is
provided in Table \ref{tab:parameters}, along with the optimal
squeezing time
$t_{\t{opt}}^{\t{OAT}}\sim N^{-2/3}\chi^{-1}\sim\eta
N^{1/3}=N^{4/3}/L$.  Lattice depths are provided in units of the
lattice recoil energy $E_R$.

@@

\section{Two-axis twisting: continuous drive}
\label{sec:continuous_drive}

We can go beyond the $\sim1/N^{2/3}$ one-axis twisting noise floor and
achieve an improved $\sim1/N$ noise floor via so-called two-axis
twisting, which can be achieved with external driving protocols.  By
turning on a clock laser resonant on the electronic transition of the
atoms, we can realize the Hamiltonian
\begin{align}
  H_{\t{drive}} = -\Omega \sum_n s_n^x + H_{\t{OAT}}^z
  = -\Omega S_x + \chi S_z^2,
\end{align}
where the drive strength $\Omega$ may generally depend on time, but
must always be much smaller in magnitude than the collective spin gap
$\eta U$ in order preserve the validity of the effective one-axis
twisting Hamiltonian (see Appendix \ref{sec:squeezing_derivation}).
Following the prescription in ref.~\cite{huang2015twoaxis} we now
modulate the drive as $\Omega\p{t}=\beta\omega\cos\p{\omega t}$ with
$\omega\gg N\chi$, and move into the rotating frame of
$-\Omega\p{t}S_x$.  After a single secular approximation which relies
only on $\omega\gg N\chi$, this procedure results in the effective
Hamiltonian
\begin{align}
  H_{\t{drive,eff}}^{\t{mod}}
  = \f{\chi}{2} \p{\sp{\J_0\p{2\beta}+1} S_z^2
    - \sp{\J_0\p{2\beta}-1} S_y^2},
\end{align}
where $\J_0$ is the zero-order Bessel function of the first kind.
Choosing a modulation index $\beta$ such that $\J_0\p{2\beta}=1/3$ or
$\J_0\p{2\beta}=-1/3$ then respectively results in the two-axis
twisting Hamiltonians
\begin{align}
  H_{\t{TAT}}^{z,x}
  = \f{\chi}{3} \p{2 S_z^2 + S_y^2}
  \simeq \f{\chi}{3} \p{S_z^2 - S_x^2},
  &&
  H_{\t{TAT}}^{y,x}
  = \f{\chi}{3} \p{S_z^2 + 2 S_y^2}
  \simeq \f{\chi}{3}\p{S_y^2 - S_x^2},
  \label{eq:H_TAT_drive}
\end{align}
where $\simeq$ denotes equality up to a global energy shift.  Given
the set of parameters in Table \ref{tab:parameters}, the two-axis
twisting Hamiltonian $H_{\t{TAT}}^{z,x}$ can be achieved with a drive
frequency $\omega=\abs{\eta U\chi}^{1/2}\approx27\times2\pi$ Hz and
modulation index $\beta\approx0.906$, and has an optimal squeezing
time of $t_{\t{opt}}^{\t{TAT}}\approx0.82$ seconds.  The figure of
merit for squeezing is the normalized minimal variance of spin in the
plane orthogonal to the mean spin vector $\bk{\v S}$:
\begin{align}
  \xi^2 \equiv \f{N}{\abs{\bk{\v S}}^2}
  \min_{\hat{\v n}\perp\bk{\v S}} \bk{\p{\v S\c\hat{\v n}}^2}
\end{align}
for $\abs{\hat{\v n}}=1$.  A comparison of this squeezing parameter
through $H_{\t{OAT}}^z$ and $H_{\t{TAT}}^{z,x}$ for the initial state
$\ket{Y}\equiv\bigotimes_n\p{\ket{\g}_n+i\ket{\e}_n}/\sqrt2$ via the
parameters in Table \ref{tab:parameters} is provided in Figure
\ref{fig:squeezing_comparison_TAT}.

\begin{figure}
  \centering \includegraphics{squeezing_comparison_TAT.pdf}
  \caption{Comparison of squeezing through $H_{\t{OAT}}^z$ and
    $H_{\t{TAT}}^{z,x}$ via the parameters in Table
    \ref{tab:parameters} and the continuous two-axis twisting protocol
    with $\omega=27\times2\pi$ Hz and $\beta=0.906$.}
  \label{fig:squeezing_comparison_TAT}
\end{figure}

@@

Two-axis twisting can also be achieved using the pulsed-drive sequence
described in ref.~\cite{liu2011spin}.  This sequence requires the
drive strength $\Omega$ much greater than the collective spin gap, or
$\abs{\Omega}\gg\eta\abs{U}$, which allows us to generate collective
spin rotations of the form $\exp\p{\pm i\theta S_x}$ with short pulses
during which we can neglect any free evolution of the atoms.  Thus
\begin{enumerate*}[label=(\roman*)]
\item acting with $\exp\sp{i\p{\pi/2}S_x}$,
\item waiting for a time $2\tau/3$,
\item acting with $\exp\sp{-i\p{\pi/2}S_x}$, and
\item waiting for a time $\tau/3$
\end{enumerate*}
realizes the unitary
\begin{align}
  U_\tau
  &= \exp\p{-i\f13\tau\chi S_z^2} \exp\p{-i\f{\pi}{2}S_x}
  \exp\p{-i\f23\tau\chi S_z^2} \exp\p{i\f{\pi}{2}S_x} \\
  &= \exp\p{-i\f13\tau\chi S_z^2} \exp\p{-i\f23\tau\chi S_y^2},
\end{align}
where if $\tau\chi\ll1$, then
\begin{align}
  U_\tau \approx \exp\sp{-i\f13\tau\chi\p{S_z^2 + 2S_y^2}}
  = \exp\p{-i\tau H_{\t{TAT}}^{y,x}}.
\end{align}
Such a sequence thus generates evolution which is equivalent to simply
evolving for a time $\tau$ under the two-axis twisting Hamiltonian
$H_{\t{TAT}}^{y,x}$.

\section{Benchmarking}

For small system sizes, we can verify the validity of the effective
one- and two-axis twisting Hamiltonians via direct simulations of a
Fermi-Hubbard (FH) model, using the spin-orbit coupling Hamiltonian
$H_{\t{SOC}}$ and interaction Hamiltonian $H_{\t{int}}$ respectively
given in \eqref{eq:H_SOC_start} and \eqref{eq:H_int_full}.  Figure
\ref{fig:squeezing_comparison_FH} shows a comparison of the squeezing
parameter $\xi^2$ after evolution of the initial state $\ket{Y}$ for
$N=8$ particles in a 1-D lattice with $L=8$ sites (and all remaining
parameters as given in Table \ref{tab:parameters}) after evolution
under
\begin{enumerate*}[label=(\roman*)]
\item the one-axis twisting Hamiltonian $H_{\t{OAT}}^z$,
\item the two-axis twisting Hamiltonian $H_{\t{OAT}}^{z,x}$,
\item free evolution of the Fermi-Hubbard model under
  $H_{\t{SOC}}+H_{\t{int}}$, and
\item driven evolution of the Fermi-Hubbard model using the
  continuous-drive protocol in Section \ref{sec:continuous_drive} with
  drive frequency $\omega=\abs{\eta U\chi}^{1/2}$ and modulation index
  $\beta=0.906$.
\end{enumerate*}
These simulation results show essentially perfect agreement between
the Fermi-Hubbard model and the one- and two-axis twisting
Hamiltonians.

\begin{figure}
  \centering \includegraphics{squeezing_comparison_FH.pdf}
  \caption{Benchmarking squeezing under the effective one- and
    two-axis Hamiltonians $H_{\t{OAT}}^z$ and $H_{\t{TAT}}^{z,x}$
    against direct simulations of the Fermi-Hubbard model with
    $N=L=8$.}
  \label{fig:squeezing_comparison_FH}
\end{figure}

% \note{[MP] Future ideas:} Make use of a cavity to enhance the
% one-axis twisting strength \cite{hu2017vacuum}?

\newpage
\appendix
